Causal analysis of COVID-19 observational data in German districts reveals effects of mobility, awareness, and temperature

### 2020-07-24 manuscript/v3_submission/data_methods.Rmd
init version for peer review submission

We constructed daily weather from four variables ("temperature", "rainfall", "humidity", "wind"). Weather data was downloaded from Deutscher Wetterdienst (DWD, [@dwd_weather]) for all weather stations in Germany below 1000 meters altitude with daily records in 2020 until 8 July 2020. District level daily weather data was aggregated per district by averaging the data from the three nearest weather stations (which includes weather stations inside the district). Missing values were imputed with mean values ($n=59$ for wind).

>>>>

We constructed daily weather from four variables ("temperature", "rainfall", "humidity", "wind"). Weather data was downloaded from Deutscher Wetterdienst (DWD, [@dwd_weather]) for all weather stations in Germany below 1000 meters altitude with daily records for our observation period. District level daily weather data was aggregated per district by averaging the data from the three nearest weather stations (which includes weather stations inside the district). Missing values were imputed with mean values ($n=59$ for wind).

### 2020-07-24 manuscript/v3_submission/discussion.Rmd
init version for peer review submission

Our objective was to identify causal effects for COVID-19 cases. We found that weather affects the reported number of infections, especially temperature (which has a reducing effect on case numbers) and rainfall (which increases case numbers). We saw that reports of high case numbers in districts led to a reduction in new infection numbers, which indicates risk-averse awareness in the population. The overall effect of mobility showed no consistent effect, however, in specific areas significant causal effects could be measured: Increasing activity in retail and recreational areas increased reported case numbers, while increased movement for essential shopping (grocery and pharmacy) led to reduced case numbers.

@@

Most research on determinants affecting case numbers of COVID-19 is restricted to single aspects [@fowler_effect_2020;@li_retrospective_2020;@shi_impact_2020;@Wang2020temperature]. To reliably identify causal drivers, one needs to adjust for confounders. To this end, we used an integrated model with variables from different aspects like mobility, awareness, weather, or socio-demographics and identified confounders by causal analysis with a directed acyclic graph. A causal approach is used in another current COVID-19 analysis [@gencoglu2020causal]. There, however, they identify the causal relationships (reconstruct a DAG), while we estimated causal effects for a given hypothesized DAG.

@@

We observed overdispersion and a substantial increase in model performance with a negative binomial regression compared to Poisson regression, which is in line with the results on COVID-19 daily case counts of @kraemer_effect_2020 and others [@liu_impact_2020;@bannister-tyrrell_preliminary_2020;@Islamm2743]. We did not model case counts with a differential equation model like the classic SIR-model [@kermack1991contributions] and its successors, since these are more suited to prediction [e.g. @rkimodellierung] while our choice of a negative binomial regression framework allowed us to estimate the effects of confounders more reliably. There are more advanced statistical methods for count data, e.g. zero-inflated models and mixed models. We tested both approaches as extensions to the negative binomial regression and experienced numerical problems and increased computing time, along with an insubstantial increase in model performance. Furthermore, our model assumed that all variables have effects proportional to the size of their measurements. It is possible that some variables show saturation effects or opposite effects for lower, medium, or high values. This could be modelled with polynomial or other transformations of the variables, which we did not employ due to limited temporal and spatial data availability. Use of a fixed DAG with effect estimation via regression assumes that data was generated by the same underlying process for the observation period. By inclusion of the successive mitigation policies as binary variables we were able to explain some of the variance caused by the changing dynamics of case numbers (similar to [@Islamm2743]).

@@

The causal effects of awareness measured via searches for "corona" and the COVID-19 burden are harder to interpret. We assume that within our model, the searches for "corona" are an insufficient proxy for awareness, while the decreasing effect for future case numbers of high daily COVID-19 burden numbers indicates it affects individual risk-behavior as a deterrent.

Similarly, the effects of temperature and rainfall can be interpreted as causal effects for indoor and outdoor activities, such that higher temperatures and low rainfall indicate more people spending time outdoor while lower temperatures and high rainfall result in indoor activities, which lead to more infections. In this light, we advocate for precautious measures like increased hygiene, face masks, and air ventilation for unavoidable indoor activities.

>>>>

Our objective was to identify causal effects for COVID-19 cases. We found that weather affects the reported number of infections, especially temperature (which has a reducing effect on case numbers) and rainfall (which increases case numbers). We saw that reports of high case numbers in districts led to a reduction in new infection numbers, which indicates risk-averse awareness in the population and/or effective public health measures to suppress a local outbreak. The overall effect of mobility showed no consistent effect, however, in specific areas significant causal effects could be measured: Increasing activity in retail and recreational areas increased reported case numbers, while increased movement for essential shopping (grocery and pharmacy) led to reduced case numbers.

@@

Most research on determinants affecting case numbers of COVID-19 is restricted to single aspects [@fowler_effect_2020;@li_retrospective_2020;@shi_impact_2020;@Wang2020temperature]. To reliably identify causal drivers, one must adjust for confounders. To this end, we used an integrated model with variables from different aspects like mobility, awareness, weather, or socio-demographics and identified confounders by causal analysis with a directed acyclic graph. A causal approach is used in another current COVID-19 analysis [@gencoglu2020causal]. There, however, they identify the causal relationships (reconstruct a DAG), while we estimated causal effects for a given hypothesized DAG.

@@

We observed overdispersion and a substantial increase in model performance with a negative binomial regression compared to Poisson regression, which is in line with the results on COVID-19 daily case counts of @kraemer_effect_2020 and others [@liu_impact_2020;@bannister-tyrrell_preliminary_2020;@Islamm2743]. We did not model case counts with a differential equation model like the classic SIR-model [@kermack1991contributions] and its successors, since these are more suited to prediction [e.g. @rkimodellierung] while our choice of a negative binomial regression framework allowed us to estimate the effects of confounders more reliably. There are more advanced statistical methods for count data, e.g. zero-inflated models and mixed models. We tested both approaches as extensions to the negative binomial regression and experienced numerical problems and increased computing time, along with an insubstantial increase in model performance. Furthermore, our model assumed that all variables have effects proportional to the size of their measurements. It is possible that some variables show saturation effects or opposite effects for low, medium, or high values. This could be modelled with polynomial or other transformations of the variables, which we did not employ due to limited temporal and spatial data availability. Use of a fixed DAG with effect estimation via regression assumes that data was generated by the same underlying process for the observation period. By inclusion of the successive mitigation policies as binary variables we were able to explain some of the variance caused by the changing dynamics of case numbers (similar to [@Islamm2743]).

@@

The causal effects of awareness measured via searches for "corona" and the COVID-19 burden are harder to interpret. We assume that within our model, the searches for "corona" are an insufficient proxy for awareness, while the decreasing effect for future case numbers of high daily COVID-19 burden indicates it affects individual risk-behavior and entails effective non-pharmaceutical interventions.

Similarly, the effects of temperature and rainfall can be interpreted as causal effects for indoor and outdoor activities, such that higher temperatures and low rainfall indicate more people spending time outdoor while lower temperatures and high rainfall result in indoor activities, which lead to more infections. Current research suggests this to be due to the prevalent airborne and respiratory droplets and aerosol transmission of the SARS-CoV-2 virus [@world2020transmission]. In this light, we advocate for precautious measures like increased hygiene, face masks, and air ventilation for unavoidable indoor activities.

### 2020-07-24 manuscript/v3_submission/manuscript.tex
init version for peer review submission

We constructed daily weather from four variables (``temperature'',
``rainfall'', ``humidity'', ``wind''). Weather data was downloaded from
Deutscher Wetterdienst (DWD, \citep{dwd_weather}) for all weather
stations in Germany below 1000 meters altitude with daily records in
2020 until 8 July 2020. District level daily weather data was aggregated
per district by averaging the data from the three nearest weather
stations (which includes weather stations inside the district). Missing
values were imputed with mean values (\(n=59\) for wind).

@@

Within our framework, we saw strong effects for mobility in
retail/recreational areas and essential shopping (grocery and pharmacy).
In the former, an increase of 1 percent point mobility compared to the
reference period (03 January to 06 February 2020) leads to an increase
of the daily reported case number by about 0.8 percent. Contrarily, a
corresponding increase of 1 percent point for the areas of
grocery/pharmacy leads to a decrease in the reported case number by
approximately 0.5 percent. Mobility on workplaces showed a small effect
of 0.3 increase in case numbers for every 1 percent point increase in
mobility. Other causal effects of mobility were insubstantial and not
consistent in their direction (99 percent confidence intervals of
estimates include 1). Figure \ref{fig:effectsrange} shows the effects of
mobility on a range of possible values. Thus, we expect an increase of
daily cases by approximately 23 percent if mobility in retail/recreation
reaches baseline levels of 0 percent difference to the reference period.
On the other hand, an increase of mobility for grocery/pharmacy by 10
percent points compared to the reference period leads to a reduction of
the infection rate by approximately 7 percent.

``Awareness'' had two opposite effects on the outcome in our DAG.
Awareness measured by Google searches for \emph{corona} had a positive
effect on the number of reported cases. An one percent point increase of
the state's Google searches (relative to other states and the
observation period) leads to an increase of approximately 1 percent. For
example, if a district shows 10 percent points more relative searches
for \emph{corona} then another one, we expect approximately 11 percent
more infections for this district after 8 days. \emph{COVID-19 burden}
(reported number of cases on day of exposure) affected the outcome
negatively, where every additional daily case in the district leads to a
0.6 percent decrease in newly reported case numbers. The last plot in
Figure \ref{fig:effectsrange} visualizes this relationship: For a local
outbreak with 20 daily cases as COVID-19 burden, we estimate as total
causal effect a subsequent reduction of infection rate by 9 percent.

Within our model, we observed a causal effect of temperature and all
other weather variables. Every increase of 1 degree Celsius in
temperature leads to a reduction of the daily reported case numbers by
approximately 0.8 percent. Similarly strong was the effect of rainfall:
One millimeter (=1 liter per square meter) more rainfall leads to an
increase of reported case numbers by approximately 1.1 percent. We
observe small effects for humidity and wind as well (higher humidity
leading to fewer cases, stronger wind leading to less cases). In
perspective (Figure \ref{fig:effectsrange}), with temperature we expect
an increase by approximately 9 percent at a daily average temperature of
\(0^{\circ}C\). For rainfall, we expect on a very rainy day with 10 mm
rainfall a corresponding increase of the infection rate by approximately
2 percent.

In all cases we opted to use the reduced optimal adjustment set over the
minimal adjustment sets because of higher pseudo-\(R^2\) values (mostly
above 0.3), except for mobility, where the minimal adjustment set had a
higher pseudo-\(R^2\). Notably, these sets always include most of our
socio-demographic variables as confounders (cf.~Table \ref{tab:adjsets})
as well as the policy variables (except for COVID-19 burden). We also
decided for the lag of 8 days based on the highest pseudo-\(R^2\) values
compared to other lags on the chosen adjustment sets. Similarly,
negative binomial regression was chosen over Poisson regression, because
the latter showed overdispersion and overall lower pseudo-\(R^2\)
values.

@@

Our objective was to identify causal effects for COVID-19 cases. We
found that weather affects the reported number of infections, especially
temperature (which has a reducing effect on case numbers) and rainfall
(which increases case numbers). We saw that reports of high case numbers
in districts led to a reduction in new infection numbers, which
indicates risk-averse awareness in the population. The overall effect of
mobility showed no consistent effect, however, in specific areas
significant causal effects could be measured: Increasing activity in
retail and recreational areas increased reported case numbers, while
increased movement for essential shopping (grocery and pharmacy) led to
reduced case numbers.

Furthermore, we made a strong case for the use of causal DAGs in
epidemiology and a pandemic like COVID-19: DAGs allow to choose
confounders for the analysis in a principled and statistically correct
way while reducing possible causes for bias. Also, the DAG formalization
allows for discussion about the underlying causal structure.

@@

Most research on determinants affecting case numbers of COVID-19 is
restricted to single aspects
\citep{fowler_effect_2020, li_retrospective_2020, shi_impact_2020, Wang2020temperature}.
To reliably identify causal drivers, one needs to adjust for
confounders. To this end, we used an integrated model with variables
from different aspects like mobility, awareness, weather, or
socio-demographics and identified confounders by causal analysis with a
directed acyclic graph. A causal approach is used in another current
COVID-19 analysis \citep{gencoglu2020causal}. There, however, they
identify the causal relationships (reconstruct a DAG), while we
estimated causal effects for a given hypothesized DAG.

@@

In addition, in alignment with other recent published studies, our
results confirm evidence which associated a negative effect of
temperature on new COVID-19 cases
\citep{bannister-tyrrell_preliminary_2020, demongeot_temperature_2020, liu_impact_2020, qi_covid-19_2020, shi_impact_2020, sobral_association_2020, tosepu_correlation_2020, Wang2020temperature, wu_effects_2020}.
It is however controversial to other scientific literature describing no
effects
\citep{briz-redon_spatio-temporal_2020, iqbal_nexus_2020, jahangiri_sensitivity_2020, juni_impact_2020, yao_no_2020}
or even converse correlations
\citep{auler_evidence_2020, xie_association_2020}. The conflicting
results might be explained by different climates and characteristics of
the populations under study. While we are confident that our strict
causal analysis resulted in effect estimates as undistorted as possible,
there might be unconsidered bias in those other studies. Further
research needs to be done to elucidate the biological characteristics of
the novel virus SARS-CoV-2 regarding its ambient temperature survival
and transmission. In regards of humidity, our findings (higher humidity
leading to increased case numbers) are in agreement with previous
research \citep{qi_covid-19_2020, wu_effects_2020, juni_impact_2020}.
Finally, we found a positive causal effect of increment precipitation
and a raise in COVID-19 cases, which supports previous observations
\citep{sobral_association_2020}.

@@

While use of a causal DAG is itself a strong tool to identify causal
effects (and not just statistical associations), it introduces two
limitations: causal assumptions within the graph (depicted by edges)
need to be well justified, and the statistical regression model that
calculates total causal effects needs to be appropriate for the task at
hand. We endorse our graph as a basis for discussion on residual
confounding.

We observed overdispersion and a substantial increase in model
performance with a negative binomial regression compared to Poisson
regression, which is in line with the results on COVID-19 daily case
counts of \citet{kraemer_effect_2020} and others
\citep{liu_impact_2020, bannister-tyrrell_preliminary_2020}. We did not
model case counts with a differential equation model like the classic
SIR-model \citep{kermack1991contributions} and its successors, since
these are more suited to prediction \citep[e.g.][]{rkimodellierung}
while our choice of a negative binomial regression framework allowed us
to estimate the effects of confounders more reliably. There are more
advanced statistical methods for count data, e.g.~zero-inflated models
and mixed models. We tested both approaches as extensions to the
negative binomial regression and experienced numerical problems and
increased computing time, along with an insubstantial increase in model
performance. Furthermore, our model assumed that all variables have
effects proportional to the size of their measurements. It is possible
that some variables show saturation effects or opposite effects for
lower, medium, or high values. This could be modelled with polynomial or
other transformations of the variables, which we did not employ due to
limited temporal and spatial data availability.

As we were using administrative data for our analysis, the results are
susceptible to the Modifiable Area Unit Problem (MAUP)
\citep{Openshaw1984}. The MAUP postulates that different regional
aggregations of the units of observation may lead to different results
and conclusions. Due to limited available data for the different
variables, there is currently no way to overcome this.

@@

In our analysis, the adverse effect of mobility in retail/recreation and
the favorable effect of mobility in grocery/pharmacy make a strong case
for the successful implementation of policies like contact restrictions
which limit the number of individual interactions that lead to
infections. This is due to retail/recreational areas encompassing mostly
places of social gatherings like restaurants and bars, malls, sports and
music venues, among others, while if people are doing more of their
essential shopping at supermarkets, they will most likely stay at home
with less contact to other people.

The causal effects of awareness measured via searches for ``corona'' and
the COVID-19 burden are harder to interpret. We assume that within our
model, the searches for ``corona'' are an insufficient proxy for
awareness, while the decreasing effect for future case numbers of high
daily COVID-19 burden numbers indicates it affects individual
risk-behavior as a deterrent.

Similarly, the effects of temperature and rainfall can be interpreted as
causal effects for indoor and outdoor activities, such that higher
temperatures and low rainfall indicate more people spending time outdoor
while lower temperatures and high rainfall result in indoor activities,
which lead to more infections. In this light, we advocate for
precautious measures like increased hygiene, face masks, and air
ventilation for unavoidable indoor activities.

@@

>>>>

We constructed daily weather from four variables (``temperature'',
``rainfall'', ``humidity'', ``wind''). Weather data was downloaded from
Deutscher Wetterdienst (DWD, \citep{dwd_weather}) for all weather
stations in Germany below 1000 meters altitude with daily records for
our observation period. District level daily weather data was aggregated
per district by averaging the data from the three nearest weather
stations (which includes weather stations inside the district). Missing
values were imputed with mean values (\(n=59\) for wind).

@@

Within our framework, we saw significant effects for mobility in
retail/recreational areas and essential shopping (grocery and pharmacy).
For retail/recreation, an increase of 1 percent point mobility compared
to the reference period (03 January to 06 February 2020) leads to an
increase of the daily reported case number by about 0.8 percent.
Contrarily, a corresponding increase of 1 percent point for the areas of
grocery/pharmacy leads to a decrease in the reported case number by
approximately 0.5 percent. Mobility on workplaces showed a small effect
of 0.3 increase in case numbers for every 1 percent point increase in
mobility. Other causal effects of mobility were insubstantial and not
consistent in their direction (99 percent confidence intervals of
estimates include 1). Figure \ref{fig:effectsrange} shows the effects of
mobility on a range of possible values. Thus, we expect an increase of
daily cases by approximately 23 percent if mobility in retail/recreation
reaches baseline levels of 0 percent difference to the reference period.
On the other hand, an increase of mobility for grocery/pharmacy by 10
percent points compared to the reference period leads to a reduction of
the infection rate by approximately 7 percent.

``Awareness'' had two opposite effects on the outcome in our DAG.
Awareness measured by Google searches for \emph{corona} had a positive
effect on the number of reported cases. An one percent point increase of
the state's Google searches (relative to other states and the
observation period) leads to an increase of approximately 1 percent. For
example, if a district shows 10 percent points more relative searches
for \emph{corona} than another one, we expect approximately 11 percent
more infections for this district after 8 days. \emph{COVID-19 burden}
(reported number of cases on day of exposure) affected the outcome
negatively, where every additional daily case in the district leads to a
0.6 percent decrease in newly reported case numbers. The last plot in
Figure \ref{fig:effectsrange} visualizes this relationship: For a local
outbreak with 25 daily cases as COVID-19 burden, we estimate as total
causal effect a subsequent reduction of infection rate by 11.8 percent.

Within our model, we observed a causal effect of temperature and all
other weather variables. Every increase of 1 degree Celsius in
temperature leads to a reduction of the daily reported case numbers by
approximately 0.8 percent. On the other hand, we found an increasing
effect of rainfall: One millimeter (=1 liter per square meter) more
rainfall leads to an increase of reported case numbers by approximately
1.1 percent. We observe effects for humidity and wind as well (higher
humidity leading to more cases, stronger wind leading to less cases). In
perspective (Figure \ref{fig:effectsrange}), with temperature we expect
an increase by approximately 9.2 percent at a daily average temperature
of \(0^{\circ}C\). For rainfall, we expect on a rainy day with 10 mm
rainfall a corresponding increase of the infection rate by approximately
9.2 percent.

In all cases we opted to use the reduced optimal adjustment set over the
minimal adjustment sets because of higher pseudo-\(R^2\) values (mostly
above 0.3), except for mobility, where the minimal adjustment set had a
higher pseudo-\(R^2\). Notably, these sets always include most of our
socio-demographic variables as confounders as well as the policy
variables (cf.~Table \ref{tab:adjsets}, with the exception being
COVID-19 burden). We also decided for the lag of 8 days based on the
highest pseudo-\(R^2\) values compared to other lags on the chosen
adjustment sets. Similarly, negative binomial regression was chosen over
Poisson regression, because the latter showed overdispersion and overall
lower pseudo-\(R^2\) values.

@@

Our objective was to identify causal effects for COVID-19 cases. We
found that weather affects the reported number of infections, especially
temperature (which has a reducing effect on case numbers) and rainfall
(which increases case numbers). We saw that reports of high case numbers
in districts led to a reduction in new infection numbers, which
indicates risk-averse awareness in the population and/or effective
public health measures to suppress a local outbreak. The overall effect
of mobility showed no consistent effect, however, in specific areas
significant causal effects could be measured: Increasing activity in
retail and recreational areas increased reported case numbers, while
increased movement for essential shopping (grocery and pharmacy) led to
reduced case numbers.

Furthermore, we made a strong case for the use of causal DAGs in
epidemiology and a pandemic like COVID-19: DAGs allow to choose
confounders for the analysis in a principled and statistically correct
way while reducing possible causes for bias. Also, the DAG formalization
allows for discussion about the underlying causal assumptions.

@@

Most research on determinants affecting case numbers of COVID-19 is
restricted to single aspects
\citep{fowler_effect_2020, li_retrospective_2020, shi_impact_2020, Wang2020temperature}.
To reliably identify causal drivers, one must adjust for confounders. To
this end, we used an integrated model with variables from different
aspects like mobility, awareness, weather, or socio-demographics and
identified confounders by causal analysis with a directed acyclic graph.
A causal approach is used in another current COVID-19 analysis
\citep{gencoglu2020causal}. There, however, they identify the causal
relationships (reconstruct a DAG), while we estimated causal effects for
a given hypothesized DAG.

@@

In addition, in alignment with other recent published studies, our
results confirm evidence which associated a negative effect of
temperature on new COVID-19 cases
\citep{bannister-tyrrell_preliminary_2020, demongeot_temperature_2020, liu_impact_2020, qi_covid-19_2020, shi_impact_2020, sobral_association_2020, tosepu_correlation_2020, Wang2020temperature, wu_effects_2020}.
It is however controversial to other scientific literature describing no
effects
\citep{briz-redon_spatio-temporal_2020, iqbal_nexus_2020, jahangiri_sensitivity_2020, juni_impact_2020, yao_no_2020}
or even converse correlations
\citep{auler_evidence_2020, xie_association_2020}. The conflicting
results might be explained by different climates and characteristics of
the populations under study. While we are confident that our strict
causal analysis resulted in effect estimates as undistorted as possible,
there might be unconsidered bias in those other studies. Further
research needs to be done to elucidate the biological characteristics of
the novel virus SARS-CoV-2 regarding its ambient temperature survival
and transmission. Finally, we found a positive causal effect of
increment precipitation and a raise in COVID-19 cases, which supports
previous observations \citep{sobral_association_2020}.

@@

While use of a causal DAG is itself a strong tool to identify causal
effects (and not just statistical associations), it introduces two
limitations: causal assumptions within the graph (depicted by edges)
need to be well justified, and the statistical regression model that
calculates total causal effects needs to be appropriate for the task at
hand. We endorse our graph as a basis for discussion on residual
confounding. We did not try to construct the DAG from the available data
(cf. \citep{gencoglu2020causal}). As such, our proposed DAG is not
entirely consistent with the data and there are conditional dependencies
between variables that cannot be dissolved by adding edges to the DAG
(e.g.~between the policies like contact restrictions and mandatory face
masks). Another way to identify potential problems in the proposed DAG
is to perform a sensitivity analysis of its structure by inspecting its
maximal ancestral graph (MAG) or its Markov equivalence class
represented by a complete partially DAG (CPDAG) and the existence of
valid adjustment sets for these generalized graphs
\citep{perkovic2017complete}. For the MAG derived from our DAG, only the
effects for exposures mobility and searches for corona can be estimated
with valid adjustment sets, while for the Markov equivalence class all
exposures but COVID-19 burden lead to valid adjustments sets. A further
analysis of these implications is out of the scope of this paper.

We observed overdispersion and a substantial increase in model
performance with a negative binomial regression compared to Poisson
regression, which is in line with the results on COVID-19 daily case
counts of \citet{kraemer_effect_2020} and others
\citep{liu_impact_2020, bannister-tyrrell_preliminary_2020, Islamm2743}.
We did not model case counts with a differential equation model like the
classic SIR-model \citep{kermack1991contributions} and its successors,
since these are more suited to prediction
\citep[e.g.][]{rkimodellierung} while our choice of a negative binomial
regression framework allowed us to estimate the effects of confounders
more reliably. There are more advanced statistical methods for count
data, e.g.~zero-inflated models and mixed models. We tested both
approaches as extensions to the negative binomial regression and
experienced numerical problems and increased computing time, along with
an insubstantial increase in model performance. Furthermore, our model
assumed that all variables have effects proportional to the size of
their measurements. It is possible that some variables show saturation
effects or opposite effects for low, medium, or high values. This could
be modelled with polynomial or other transformations of the variables,
which we did not employ due to limited temporal and spatial data
availability. Use of a fixed DAG with effect estimation via regression
assumes that data was generated by the same underlying process for the
observation period. By inclusion of the successive mitigation policies
as binary variables we were able to explain some of the variance caused
by the changing dynamics of case numbers (similar to
\citep{Islamm2743}).

We stress the point that our effects were deduced on an aggregate
(district) level in the absence of available data on an individual
level. As such, conclusions about effects cannot be transferred on
individuals without the possibility for an ecological fallacy.
Furthermore, as we were using administrative data for our analysis, the
results are susceptible to the Modifiable Area Unit Problem (MAUP)
\citep{Openshaw1984}. The MAUP postulates that different regional
aggregations of the units of observation may lead to different results
and conclusions. Due to limited available data for the different
variables, there is currently no way to overcome these problems that are
inherent to all analyses on aggregated data level.

@@

In our analysis, the adverse effect of mobility in retail/recreation and
the favorable effect of mobility in grocery/pharmacy indicate that
policies like contact restrictions which limit the number of individual
interactions can lead to reduced infection numbers. This is due to
retail/recreational areas encompassing mostly places of social
gatherings like restaurants and bars, malls, sports and music venues,
among others, while if people are doing more of their essential shopping
at supermarkets, they will most likely stay at home with less contact to
other people.

The causal effects of awareness measured via searches for ``corona'' and
the COVID-19 burden are harder to interpret. We assume that within our
model, the searches for ``corona'' are an insufficient proxy for
awareness, while the decreasing effect for future case numbers of high
daily COVID-19 burden indicates it affects individual risk-behavior and
entails effective non-pharmaceutical interventions.

Similarly, the effects of temperature and rainfall can be interpreted as
causal effects for indoor and outdoor activities, such that higher
temperatures and low rainfall indicate more people spending time outdoor
while lower temperatures and high rainfall result in indoor activities,
which lead to more infections. Current research suggests this to be due
to the prevalent airborne and respiratory droplets and aerosol
transmission of the SARS-CoV-2 virus \citep{world2020transmission}. In
this light, we advocate for precautious measures like increased hygiene,
face masks, and air ventilation for unavoidable indoor activities.

@@

\section*{Acknowledgments}\label{acknowledgments}
\addcontentsline{toc}{section}{Acknowledgments}

We are thankful for feedback from Thomas Czihal, Johannes Textor, and
Ralph Brinks, who gave helpful suggestions on an earlier draft of the
manuscript.
